---
title: 'Project 6: Randomization and Matching'
output: pdf_document
---

# Introduction

In this project, you will explore the question of whether college education causally affects political participation. Specifically, you will use replication data from \href{https://papers.ssrn.com/sol3/papers.cfm?abstract_id=1409483}{Who Matches? Propensity Scores and Bias in the Causal Eﬀects of Education on Participation} by former Berkeley PhD students John Henderson and Sara Chatfield. Their paper is itself a replication study of \href{https://www.jstor.org/stable/10.1017/s0022381608080651}{Reconsidering the Effects of Education on Political Participation} by Cindy Kam and Carl Palmer. In their original 2008 study, Kam and Palmer argue that college education has no effect on later political participation, and use the propensity score matching to show that pre-college political activity drives selection into college and later political participation. Henderson and Chatfield in their 2011 paper argue that the use of the propensity score matching in this context is inappropriate because of the bias that arises from small changes in the choice of variables used to model the propensity score. They use \href{http://sekhon.berkeley.edu/papers/GenMatch.pdf}{genetic matching} (at that point a new method), which uses an approach similar to optimal matching to optimize Mahalanobis distance weights. Even with genetic matching, they find that balance remains elusive however, thus leaving open the question of whether education causes political participation.

You will use these data and debates to investigate the benefits and pitfalls associated with matching methods. Replication code for these papers is available online, but as you'll see, a lot has changed in the last decade or so of data science! Throughout the assignment, use tools we introduced in lab from the \href{https://www.tidyverse.org/}{tidyverse} and the \href{https://cran.r-project.org/web/packages/MatchIt/MatchIt.pdf}{MatchIt} packages. Specifically, try to use dplyr, tidyr, purrr, stringr, and ggplot instead of base R functions. While there are other matching software libraries available, MatchIt tends to be the most up to date and allows for consistent syntax.

# Data

The data is drawn from the \href{https://www.icpsr.umich.edu/web/ICPSR/studies/4023/datadocumentation#}{Youth-Parent Socialization Panel Study} which asked students and parents a variety of questions about their political participation. This survey was conducted in several waves. The first wave was in 1965 and established the baseline pre-treatment covariates. The treatment is whether the student attended college between 1965 and 1973 (the time when the next survey wave was administered). The outcome is an index that calculates the number of political activities the student engaged in after 1965. Specifically, the key variables in this study are:

\begin{itemize}
    \item \textbf{college}: Treatment of whether the student attended college or not. 1 if the student attended college between 1965 and 1973, 0 otherwise.
    \item \textbf{ppnscal}: Outcome variable measuring the number of political activities the student participated in. Additive combination of whether the student voted in 1972 or 1980 (student\_vote), attended a campaign rally or meeting (student\_meeting), wore a campaign button (student\_button), donated money to a campaign (student\_money), communicated with an elected official (student\_communicate), attended a demonstration or protest (student\_demonstrate), was involved with a local community event (student\_community), or some other political participation (student\_other)
\end{itemize}

Otherwise, we also have covariates measured for survey responses to various questions about political attitudes. We have covariates measured for the students in the baseline year, covariates for their parents in the baseline year, and covariates from follow-up surveys. \textbf{Be careful here}. In general, post-treatment covariates will be clear from the name (i.e. student\_1973Married indicates whether the student was married in the 1973 survey). Be mindful that the baseline covariates were all measured in 1965, the treatment occurred between 1965 and 1973, and the outcomes are from 1973 and beyond. We will distribute the Appendix from Henderson and Chatfield that describes the covariates they used, but please reach out with any questions if you have questions about what a particular variable means.

```{r}
# Load tidyverse and MatchIt
# Feel free to load other libraries as you wish
library(tidyverse)
library(MatchIt)
library(MatchIt)
library(ggplot2)
library(dplyr)
library(tableone)
library(cobalt)
library(gridExtra) 
library(optmatch)

# Load ypsps data
ypsps <- read_csv('data/ypsps.csv')
head(ypsps)
glimpse(ypsps)
```

# Randomization

Matching is usually used in observational studies to to approximate random assignment to treatment. But could it be useful even in randomized studies? To explore the question do the following:

\begin{enumerate}
    \item Generate a vector that randomly assigns each unit to either treatment or control
    \item Choose a baseline covariate (for either the student or parent). A binary covariate is probably best for this exercise.
    \item Visualize the distribution of the covariate by treatment/control condition. Are treatment and control balanced on this covariate?
    \item Simulate the first 3 steps 10,000 times and visualize the distribution of treatment/control balance across the simulations.
\end{enumerate}

```{r}

# Generate a vector that randomly assigns each unit to treatment/control
library(dplyr)

set.seed(10)  # reproducibility
ypsps <- ypsps %>%
  mutate(treatment = sample(c(0, 1), n(), replace = TRUE))  #O=control, 1=treatment


# Choose a baseline covariate (use dplyr for this)
# Choosing gender even though famously not a binary lol.
# It says in the assignment to be careful about the covariates but that's impossible
# without more context -- I can't even figure out for sure whether 0 is male or not. I'm going to 
# proceed with that assumption because that's kinda standard

library(ggplot2)

ypsps %>%
  mutate(student_Gen = ifelse(student_Gen == 0, "Male", "Female"))

# Plot without ingo
ggplot(ypsps, aes(x = student_Gen)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Student Gender", x = "Gender", y = "Count") +
  theme_minimal()

# Plotting with my guess as to what 0 and 1 mean
ggplot(ypsps %>%
         mutate(student_Gen = ifelse(student_Gen == 0, "Male", "Female")), 
       aes(x = student_Gen)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Student Gender", x = "Gender", y = "Count") +
  theme_minimal()

# Visualize the distribution by treatment/control (ggplot)
ggplot(ypsps %>%
         mutate(student_Gen = ifelse(student_Gen == 0, "Female", "Male")), 
       aes(x = student_Gen)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Student Gender by Treatment/Control", x = "Gender", y = "Count") +
  facet_wrap(~ treatment, labeller = labeller(treatment = c("0" = "Control", "1" = "Treatment"))) +
  theme_minimal()

# trying to add data labels
ypsps %>%
  mutate(student_Gen = ifelse(student_Gen == 0, "Female", "Male")) %>%
  group_by(treatment, student_Gen) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(treatment) %>%
  mutate(pct = count / sum(count) * 100) %>%
  ggplot(aes(x = student_Gen, y = pct, fill = student_Gen)) +
  geom_bar(stat = "identity", color = "black", show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -0.5) +
  labs(title = "Percentage of Male and Female Students by Treatment/Control", x = "Gender", y = "Percentage") +
  facet_wrap(~ treatment, labeller = labeller(treatment = c("0" = "Control", "1" = "Treatment"))) +
  theme_minimal()

#Alternately
ggplot(ypsps, aes(x = factor(treatment), fill = factor(student_Gen))) +
  geom_bar(position = "fill", color = "black") +
  labs(title = "Distribution of Student Gender by Treatment/Control",
       x = "Treatment/Control",
       y = "Proportion",
       fill = "Gender") +
  scale_x_discrete(labels = c("0" = "Control", "1" = "Treatment")) +
  theme_minimal()

#They seem pretty balanced! Both are around 50%. 

# Simulate the process 1000 times
sim_results <- replicate(1000, {
  # Randomly assign treatment/control
  temp_data <- ypsps %>%
    mutate(treatment = sample(c(0, 1), n(), replace = TRUE))
  
  # Calculate proportions of male/female in treatment and control groups
  treatment_props <- temp_data %>%
    group_by(treatment, student_Gen) %>%
    summarise(count = n(), .groups = 'drop') %>%  # Add .groups = 'drop' to avoid grouping by treatment
    group_by(treatment) %>%
    mutate(percentage = count / sum(count) * 100) %>%
    filter(student_Gen == "0") %>%
    select(treatment, percentage) %>%
    pivot_wider(names_from = treatment, values_from = percentage)  # Pivot the data
  
  # Adjust mutate based on the column names we see
  if ("1" %in% names(treatment_props) && "0" %in% names(treatment_props)) {
    treatment_props <- treatment_props %>%
      mutate(diff = `1` - `0`)  # If columns are named '1' and '0'
  } else {
    # If columns have different names like '...1' and '...0'
    treatment_props <- treatment_props %>%
      mutate(diff = `...1` - `...0`)  # Adjust according to actual column names
  }
  
  # Return the difference between treatment and control
  return(treatment_props$diff)
})

# Convert to data frame for visualization
sim_results_df <- data.frame(diff = unlist(sim_results))

# Visualize the distribution of differences in male proportions between treatment/control across simulations
ggplot(sim_results_df, aes(x = diff)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Treatment/Control Balance Across 1000 Simulations",
       x = "Difference in Gender Proportion (Treatment - Control)",
       y = "Frequency") +
  theme_minimal()


```



## Questions
\begin{enumerate}
    \item \textbf{What do you see across your simulations? Why does independence of treatment assignment and baseline covariates not guarantee balance of treatment assignment and baseline covariates?}
\end{enumerate}

Your Answer: Across 1,000 simulations, it looks like the gender is pretty balanced, but it's clear that there are available random selections that would give me highly undesirable gender balances (10-1 female or male in the treatment group, for example). This is not surprising, since, as the questions states, "independence of treatment assignment and baseline covariates not guarantee balance of treatment assignment and baseline covariates." "Independence" just means that the one doesn't have anything to do with the other; it doesn't _guarantee_ that distribution will be balanced. Randomization might just result in a skewed treatment or control group, or the sample might just not have that many men to distribute between groups.  

# Propensity Score Matching

## One Model
Select covariates that you think best represent the "true" model predicting whether a student chooses to attend college, and estimate a propensity score model to calculate the Average Treatment Effect on the Treated (ATT). Plot the balance of the top 10 (or fewer if you select fewer covariates). Report the balance of the p-scores across both the treatment and control groups, and using a threshold of standardized mean difference of p-score $\leq .1$, report the number of covariates that meet that balance threshold.

NH: Only one covariate meets the balance threshold for now!

```{r}
# Select covariates that represent the "true" model for selection, fit model

# List of covariates
covariates <- c("parent_EducHH", "parent_Employ", "student_Gen", "student_Race", "parent_Vote")  

# Select only relevant columns
ypsps_subset <- ypsps %>%
  dplyr::select(college, all_of(covariates), student_ppnscal)

# Estimate propensity scores using logistic regression
ps_model <- glm(college ~ ., data = ypsps_subset %>% select(-student_ppnscal), family = binomial())

# Predict propensity scores
ypsps_subset$p_score <- predict(ps_model, type = "response")

# Plot the balance for the top 10 covariates

# exclude treatment & outcome variables
covariates <- setdiff(names(ypsps_subset), c("college", "student_ppnscal", "p_score"))

# Create a table
table <- CreateTableOne(vars = covariates, 
                        strata = "college", 
                        data = ypsps_subset, 
                        test = FALSE)

# Extract SMD values
smd_values <- print(table, smd = TRUE) %>% as.data.frame()

smd_values$SMD <- as.numeric(trimws(smd_values$SMD))

smd_values <- smd_values %>% filter(!is.na(SMD))

covariates_with_balance <- sum(smd_values$SMD <= 0.1, na.rm = TRUE)

total_covariates <- nrow(smd_values)

prop_balance <- covariates_with_balance / total_covariates


cat("Total number of covariates:", total_covariates, "\n")
cat("Number of covariates meeting the balance threshold (SMD ≤ 0.1):", covariates_with_balance, "\n")
cat("Proportion of covariates meeting the balance threshold:", round(prop_balance, 3), "\n")


# Select top 10 most imbalanced covariates (highest SMD)
# 4/1/25: I've asked for clarity, but I'm not sure what the variables are for now. Leaving these ones here for now, but I'll refine if I get more information before I have to leave
top_10_smd <- smd_values %>%
  arrange(desc(SMD)) %>%
  slice(1:10) %>%
  mutate(Covariate = rownames(.))

# Report the overall balance and the proportion of covariates that meet the balance threshold
ggplot(top_10_smd, aes(x = SMD, y = reorder(Covariate, SMD))) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_vline(xintercept = 0.1, color = "red", linetype = "dashed", size = 1.2) +
  labs(title = "Covariate Balance Before Matching",
       x = "Standardized Mean Difference",
       y = "Covariate") +
  theme_minimal()
```

## Simulations

Henderson/Chatfield argue that an improperly specified propensity score model can actually \textit{increase} the bias of the estimate. To demonstrate this, they simulate 800,000 different propensity score models by choosing different permutations of covariates. To investigate their claim, do the following:

\begin{itemize}
    \item Using as many simulations as is feasible (at least 10,000 should be ok, more is better!), randomly select the number of and the choice of covariates for the propensity score model.
    \item For each run, store the ATT, the proportion of covariates that meet the standardized mean difference $\leq .1$ threshold, and the mean percent improvement in the standardized mean difference. You may also wish to store the entire models in a list and extract the relevant attributes as necessary.
    \item Plot all of the ATTs against all of the balanced covariate proportions. You may randomly sample or use other techniques like transparency if you run into overplotting problems. Alternatively, you may use plots other than scatterplots, so long as you explore the relationship between ATT and the proportion of covariates that meet the balance threshold.
    \item Finally choose 10 random models and plot their covariate balance plots (you may want to use a library like \href{https://cran.r-project.org/web/packages/gridExtra/index.html}{gridExtra} to arrange these)
\end{itemize}

\textbf{Note: There are lots of post-treatment covariates in this dataset (about 50!)! You need to be careful not to include these in the pre-treatment balancing. Many of you are probably used to selecting or dropping columns manually, or positionally. However, you may not always have a convenient arrangement of columns, nor is it fun to type out 50 different column names. Instead see if you can use dplyr 1.0.0 functions to programatically drop post-treatment variables (\href{https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-select-rename-relocate/}{here} is a useful tutorial).}

```{r}
#cleaning
ypsps_clean <- ypsps %>%
#  mutate(ATT = NA, balanced_proportion = NA, mean_improvement = NA)  %>%
  select(-contains("1973"), -contains("1982"))

colSums(is.na(ypsps_clean))

# I think rows with missing values are messing me up bc matchit doens't accept missing values. I'm going to just drop those rows for now, might explore other solutions later on
ypsps_clean <- na.omit(ypsps_clean)  
glimpse(ypsps_clean)
colSums(is.na(ypsps_clean))


# Quick check for balance between treatment and control
table(ypsps_clean$college)
table(ypsps_clean$student_ppnscal)

```

```{r}
# Random selection of features 

# Fit p-score models and save ATTs, proportion of balanced covariates, and mean percent balance improvement

# Plot ATT v. proportion

# 10 random covariate balance plots (hint try gridExtra)

# Note: ggplot objects are finnicky so ask for help if you're struggling to automatically create them; consider using functions!
```


```{r}
# # One run-through
# # Random selection of features 
# set.seed(10) 
# selected_covariates1 <- sample(colnames(ypsps_clean), 7)
# print(selected_covariates1)
# vars_to_keep1 <- c(selected_covariates1, "college", "student_ppnscal")
# 
# #limiting to selected covariates
# ypsps_clean1 <- ypsps_clean %>%
# #  mutate(ATT = NA, balanced_proportion = NA, mean_improvement = NA)  %>%
#   select(all_of(vars_to_keep1))
# 
# # Fit p-score models and save ATTs, proportion of balanced covariates, and mean percent balance improvement
# 
# #Fit model
# ps_model <- matchit(college ~ ., data = ypsps_clean1, method = "nearest", distance = "logit")
# 
# # Get matches
# matched_data <- match.data(ps_model)
# 
# # ATT using a regression model
# att_model <- lm(student_ppnscal ~ college, data = matched_data)
# att <- coef(att_model)["college"]
# 
# # Get balance table
# balance_table <- bal.tab(ps_model, un = TRUE, thresholds = c(m = 0.1))
# #debugging...
# # print(balance_table)
# # str(balance_table$Balance)
# # head(balance_table$Balance)
# smd_unadjusted <- as.numeric(balance_table$Balance[, "Diff.Un"])  # Before matching
# smd_values <- as.numeric(balance_table$Balance[, "Diff.Adj"])     # After matching
# 
# # Proportion of covariates meeting SMD ≤ 0.1
# prop_balance <- mean(abs(smd_values) <= 0.1, na.rm = TRUE)
# 
# # Mean percent improvement in SMD
# smd_improvement <- mean((abs(smd_unadjusted) - abs(smd_values)) / abs(smd_unadjusted) * 100, na.rm = TRUE)
# 
# results <- data.frame(
#   ATT = att,
#   Prop_Balanced = prop_balance,
#   Mean_SMD_Improvement = smd_improvement
# )
# 
# print(results)

# Plot ATT v. proportion

# 10 random covariate balance plots (hint try gridExtra)

# Note: ggplot objects are finnicky so ask for help if you're struggling to automatically create them; consider using functions!
```



```{r}
#loop attempt

# Set the seed for reproducibility
set.seed(10)

# Initialize an empty dataframe to store the results
final_results <- data.frame(
  Run = integer(0),
  ATT = numeric(0),
  Prop_Balanced = numeric(0),
  Mean_SMD_Improvement = numeric(0)
)

# Loop to run 20 times
for (i in 1:30) {
  # Randomly select 7 covariates
  selected_covariates1 <- sample(colnames(ypsps_clean), 10)
  
  # Select the covariates and outcome variables to keep
  vars_to_keep1 <- c(selected_covariates1, "college", "student_ppnscal")
  
  # Create a subset of the data with the selected variables
  ypsps_clean1 <- ypsps_clean %>%
    select(all_of(vars_to_keep1))
  
  # Fit the propensity score model
  ps_model <- matchit(college ~ ., data = ypsps_clean1, method = "nearest", distance = "logit")
  
  # Get the matched data
  matched_data <- match.data(ps_model)
  
  # Estimate ATT using a regression model
  att_model <- lm(student_ppnscal ~ college, data = matched_data)
  att <- coef(att_model)["college"]
  
  # Get the balance table
  balance_table <- bal.tab(ps_model, un = TRUE, thresholds = c(m = 0.1))
  
  # Extract the standardized mean differences (SMD)
  smd_unadjusted <- as.numeric(balance_table$Balance[, "Diff.Un"])  # Before matching
  smd_values <- as.numeric(balance_table$Balance[, "Diff.Adj"])     # After matching
  
  # Proportion of covariates meeting SMD ≤ 0.1
  prop_balance <- mean(abs(smd_values) <= 0.1, na.rm = TRUE)
  
  # Calculate Mean percent improvement in SMD
  smd_improvement <- mean((abs(smd_unadjusted) - abs(smd_values)) / abs(smd_unadjusted) * 100, na.rm = TRUE)
  
  # Store the results for the current iteration
  results <- data.frame(
    Run = i,
    ATT = att,
    Prop_Balanced = prop_balance,
    Mean_SMD_Improvement = smd_improvement
  )
  
  # Append the results to the final results dataframe
  final_results <- rbind(final_results, results)
}

# Print the final results
print(final_results)

```

```{r}
#plot

# Plot ATT vs. Proportion of Balanced Covariates
ggplot(final_results, aes(x = Prop_Balanced, y = ATT)) +
  geom_point() +  # Scatter plot
#  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add a linear regression line
  labs(
    title = "ATT vs. Proportion of Balanced Covariates for PS matching",
    x = "Proportion of Balanced Covariates",
    y = "Average Treatment Effect (ATT)"
  ) +
  theme_minimal()  # Clean theme

```
```{r}
# Set seed for reproducibility
set.seed(10)

# Initialize a list to store the plots
balance_plots <- list()

# Loop to create 5 models
for (i in 1:5) {
  # Randomly select 7 covariates for this model
  selected_covariates <- sample(colnames(ypsps_clean), 7)
  
  # Select variables to keep (add "college" and "student_ppnscal")
  vars_to_keep <- c(selected_covariates, "college", "student_ppnscal")
  
  # Create a subset of the data with selected variables
  ypsps_clean_subset <- ypsps_clean %>%
    select(all_of(vars_to_keep))
  
  # Fit the propensity score model using nearest neighbor matching
  ps_model <- matchit(college ~ ., data = ypsps_clean_subset, method = "nearest", distance = "logit")
  
  # Get balance diagnostics
  balance_table <- bal.tab(ps_model, un = TRUE, thresholds = c(m = 0.1),standardize = TRUE )
  
  # Create a balance plot for this model
  balance_plot <- love.plot(balance_table, threshold = 0.1, abs = TRUE, title = paste("Model", i))
    balance_plot <- balance_plot + 
    theme(
      axis.text.x = element_text(size = 8),   # Adjust x-axis text size
      axis.text.y = element_text(size = 4),   # Adjust y-axis text size
      axis.title.x = element_text(size = 10), # Adjust x-axis title size
      axis.title.y = element_text(size = 10), # Adjust y-axis title size
      plot.title = element_text(size = 12)    # Adjust plot title size
    ) 
    
  # Store the plot in the list
  balance_plots[[i]] <- balance_plot
}

# Use gridExtra to arrange the plots in a grid (2 columns and 3 rows)
#grid.arrange(grobs = balance_plots, ncol = 2)
grid.arrange(grobs = balance_plots, ncol = 2, heights = c(1, 1, 1), widths = rep(1, 2))
```



## Questions

\begin{enumerate}
    \item \textbf{How many simulations resulted in models with a higher proportion of balanced covariates? Do you have any concerns about this?}
    Your Answer:...
    \item \textbf{Analyze the distribution of the ATTs. Do you have any concerns about this distribution?}
    Your Answer:...
    \item \textbf{Do your 10 randomly chosen covariate balance plots produce similar numbers on the same covariates? Is it a concern if they do not?}
    Your Answer:...
\end{enumerate}

# Matching Algorithm of Your Choice

## Simulate Alternative Model

Henderson/Chatfield propose using genetic matching to learn the best weights for Mahalanobis distance matching. Choose a matching algorithm other than the propensity score (you may use genetic matching if you wish, but it is also fine to use the greedy or optimal algorithms we covered in lab instead). Repeat the same steps as specified in Section 4.2 and answer the following questions:

```{r}
#cleaning
ypsps_clean <- ypsps %>%
#  mutate(ATT = NA, balanced_proportion = NA, mean_improvement = NA)  %>%
  select(-contains("1973"), -contains("1982"))

colSums(is.na(ypsps_clean))

# I think rows with missing values are messing me up bc matchit doens't accept missing values. I'm going to just drop those rows for now, might explore other solutions later on
ypsps_clean <- na.omit(ypsps_clean)  
glimpse(ypsps_clean)
colSums(is.na(ypsps_clean))


```



```{r}
# loop for genetic matching
# It looks like these libraries are necessary for genetic matching, but earlier chunks don't seem to work if I load these libraries earlier in the code. In fact, I need to clear the environment and restart r if I load them too early.
# Update: I think there's another "select" function so I specified "dplyr" and that bug is fixed. 

library(rgenoud)

# Set the seed for reproducibility
set.seed(10)

# Initialize an empty dataframe to store the results
gen_final_results <- data.frame(
  Run = integer(0),
  ATT = numeric(0),
  Prop_Balanced = numeric(0),
  Mean_SMD_Improvement = numeric(0)
)

# Loop to run 20 times
for (i in 1:30) {
  # Randomly select 7 covariates
  selected_covariates1 <- sample(colnames(ypsps_clean), 10)
  
  # Select the covariates and outcome variables to keep
  vars_to_keep1 <- c(selected_covariates1, "college", "student_ppnscal")
  
  # Create a subset of the data with the selected variables
  ypsps_clean1 <- ypsps_clean %>%
    dplyr::select(all_of(vars_to_keep1))
  
  # Fit the propensity score model 
  greedy_model <- matchit(college ~ ., data = ypsps_clean1, method = "genetic", distance = "logit")
  
  # Get the matched data
  matched_data <- match.data(greedy_model)
  
  # Estimate ATT using a regression model
  att_model <- lm(student_ppnscal ~ college, data = matched_data)
  att <- coef(att_model)["college"]
  
  # Get the balance table
  balance_table <- bal.tab(greedy_model, un = TRUE, thresholds = c(m = 0.1))
  
  # Extract the standardized mean differences (SMD)
  smd_unadjusted <- as.numeric(balance_table$Balance[, "Diff.Un"])  # Before
  smd_values <- as.numeric(balance_table$Balance[, "Diff.Adj"])     # After 
  
  # Proportion of covariates meeting SMD ≤ 0.1
  prop_balance <- mean(abs(smd_values) <= 0.1, na.rm = TRUE)
  
  # Calculate Mean percent improvement in SMD
  smd_improvement <- mean((abs(smd_unadjusted) - abs(smd_values)) / abs(smd_unadjusted) * 100, na.rm = TRUE)
  
  # Store the results for the current iteration
  results <- data.frame(
    Run = i,
    ATT = att,
    Prop_Balanced = prop_balance,
    Mean_SMD_Improvement = smd_improvement
  )
  
  # Append the results to the final results dataframe
  gen_final_results <- rbind(gen_final_results, results)
}

# Print the final results
print(gen_final_results)

```



```{r}
#plot

# Plot ATT vs. Proportion of Balanced Covariates
ggplot(gen_final_results, aes(x = Prop_Balanced, y = ATT)) +
  geom_point() +  # Scatter plot
#  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add a linear regression line
  labs(
    title = "ATT vs. Proportion of Balanced Covariates for Genetic Matching",
    x = "Proportion of Balanced Covariates",
    y = "Average Treatment Effect (ATT)"
  ) +
  theme_minimal()  # Clean theme

```


```{r}
# Remove post-treatment covariates

# Randomly select features

# Simulate random selection of features 10k+ times

# Fit  models and save ATTs, proportion of balanced covariates, and mean percent balance improvement

# Plot ATT v. proportion

# 10 random covariate balance plots (hint try gridExtra)
# Note: ggplot objects are finnicky so ask for help if you're struggling to automatically create them; consider using functions!
```

```{r}
# Visualization for distributions of percent improvement
```

## Questions

\begin{enumerate}
    \item \textbf{Does your alternative matching method have more runs with higher proportions of balanced covariates?}
     Your Answer:...
    \item \textbf{Use a visualization to examine the change in the distribution of the percent improvement in balance in propensity score matching vs. the distribution of the percent improvement in balance in your new method. Which did better? Analyze the results in 1-2 sentences.}
    Your Answer:...
\end{enumerate}

\textbf{Optional:} Looking ahead to the discussion questions, you may choose to model the propensity score using an algorithm other than logistic regression and perform these simulations again, if you wish to explore the second discussion question further.

# Discussion Questions

\begin{enumerate}
    \item \textbf{Why might it be a good idea to do matching even if we have a randomized or as-if-random design?}
    Your Answer:...
    \item \textbf{The standard way of estimating the propensity score is using a logistic regression to estimate probability of treatment. Given what we know about the curse of dimensionality, do you think there might be advantages to using other machine learning algorithms (decision trees, bagging/boosting forests, ensembles, etc.) to estimate propensity scores instead?}
    Your Answer:...
\end{enumerate}